Open archaeology, open source?

### 2023-06-27 analysis/_01-introduction.qmd
Refining the intro and background some more.

Open archaeology constitutes a series of practices and principles that support greater transparency to research processes and access to data that underlie synthetic outcomes (Beck and Neylon 2012: 479-480). As part of the broader open science movement, open archaeology promotes more thoughtful scholarly communication practices, most notably in operations relating to publishing, data-sharing, and education, but it continues to expand its reach into the arenas of bibliographic metadata, review processes, and grantsmanship. Archaeologists are also actively venturing into open source software development as means of sharing their research processes and creating tools and resources for general use.

However, academic implementation of open source software development models is complicated by the open source movement's distinct history and its nuanced relationship with the open science movement. The open source movement developed out of pressures and circumstances that differ from those in which the open science movement operates, and yet the former serves as direct inspiration for the latter. For instance, the Open Knowledge Foundation, which publishes a widely accepted defitnion of "open" in the context of scholarly communication, draws directly from the definition of "open source" published by the Open Source Initiative (@openknowledgefoundation2023; @opensourceinitiative2007). The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plaintext, diffs, emphasis on formal licencing, the whole hacker aesthetic). However, academic work, including the development of academic software, differs signififcantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, the Firefox web browser, and openSSL. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional pressures. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers (Kelty 2008:xx).

The relationship between open science and open source is also complicated by rhetorical claims that, based on our personal experiences, are discordant with how academics actually _do_ open source. Does academic open source actually make research processes more transparent and improve research outcomes? Is it actually boosting efficiency by establishing a common store of knowledge and productive code? Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass? Despite hopeful aspirations espoused by open archaeology advocates (Kansa  et al. 2014; Kintigh et al. 2015; others), these outcomes are not a given. Based on casual observations from our own experiences, we find that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. Unfortunately, these values are not widespread among archaeologists, even those who are staunch supporters of open science, due to systemic challenges and tensions.

That being said, archaeologists are prolific software developers, as is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources, which records XXX items as of the time of publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

Our statistical and network analyses consider how archaeological software development may be benefiting from or missing out on the affordances that open source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency. We found that the vast majority of projects lack any collective effort and die shortly after they are initiated. Projects that are maintained by multiple contributors tend to be backed by funded initiatives who employ locally and socially engaged individuals, or are warranted by participation in a scholarly community exhibiting genuine need for a particular kind of resource. Moreover, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. Contrary to popular claims about open source being inherently distributed, resilient, and open-ended, we found that, overall, archaeological open source is actually quite centralized, fragile, and based nearly exclusively on existing professional connections and endeavours.

>>>>

In their seminal vision for open archaeology, Beck and Neylon (2012: 480-481) identified the movement as comprising a series of principles and practices "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity". Open archaeology therefore promotes more thoughtful scholarly communication practices, most notably in operations relating to publishing, data-sharing, education, and review processes. Archaeologists are also actively engaged in open source software development as means of sharing their research processes and creating tools and resources for general use.

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation, which publishes a widely accepted defitnion of "open" in the context of scholarly communication, explicitly refers to the definition of "open source" published by the Open Source Initiative, an authoritative open source advocacy group (@openknowledgefoundation2023; @opensourceinitiative2007). The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plaintext, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs signififcantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL, and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers (Kelty 2008:xx).

The relationship between open science and open source is also complicated by rhetorical claims that, based on our personal experiences, are discordant with how academics actually _do_ open source. Does academic open source actually make research processes more transparent and improve research outcomes? Is it actually boosting efficiency by establishing a common store of knowledge and productive code? Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass? Despite hopeful aspirations espoused by open archaeology advocates (Kansa  et al. 2014; Kintigh et al. 2015; others), these outcomes are not a given. We believe that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. However, practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.

That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes XXX records as of the time of publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

Our statistical and network analyses consider how archaeological software development may be benefiting from or missing out on the affordances that open source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency. We found that the vast majority of projects lack any collective effort and activity tends to abruptly end shortly after work is initiated. Projects that are maintained by multiple contributors tend to be backed by funded initiatives who employ locally and socially engaged individuals, or are warranted by participation in a scholarly community exhibiting genuine need for a particular kind of resource. Moreover, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. Contrary to popular claims about open source being inherently distributed, resilient, and open-ended, we found that, overall, archaeological open source is actually quite centralized, fragile, and based nearly exclusively on existing professional connections and endeavours.

### 2023-06-27 analysis/_02-background.qmd
Refining the intro and background some more.

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development (@kelty2008: chapter 3, especially starting at page 99). By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision. An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional. Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes (@coleman2012).

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants. They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience (@kelty2008; Ratto dissertation and derived works).

@@

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" (@openknowledgefoundation2023). This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scienctific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

@@

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain (cf. @tennant2020). But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (inlcuding: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed and credit has been allocated, it is common for scientists to mark their projects as completed and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly, and contributors may remain actively or sporadically involved to satisfy users' needs.

>>>>

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development (@kelty2008: chapter 3, especially starting at page 99). By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision (Kelty 2008 on the emacs saga). An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional (Kelty 2008: 220). Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes (@coleman2012; O'Neil 2009: the debian saga).

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants (O'Neil 2009). They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience (@kelty2008; Ratto dissertation and derived works).

@@

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" (@openknowledgefoundation2023). This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates (including archaeologists, as elicited by Beck and Neylon 2012; Marwick et al. 2017) claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scienctific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

@@

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain (@tennant2020). But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (inlcuding: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software.
